Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany

### 2025-06-10 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
latest paper added, figures added, citations fixed

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author: Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta, Jonathan Renkel, Caroline Arnold (order tbd)
author-running: T. Khan et al.
abstract: |
  Advances in deep learning models have opened up new avenues for computer vision tasks, like tree crown segmentations from remote sensing images. As more and more models become available, the challenges of applying the models to really large spatial scales (beyond the forest/plot level) become visible. One of these challenges is the validation of the tree crown detection and segmentation. Imprecise detections/segmentations can lead to inaccuracies in downstream tasks in forestry and urban planning, creating a barrier for the application of these models to real-life use cases. Attributes of training data, e.g tree sizes, phenological stages, number of polygon vertices, shadows, impact the prediction accuracy of the segmentation models, hence training data should be chosen with due diligence. Here we identify some of the key validation challenges based on the collective experiences of not just training deep learning models, but also applying the models to very large spatial scales in Germany (federal state level).
keywords: [First Keyword, Second Keyword, Another Keyword.]
---

## Introduction

Accurately mapping tree crowns over large areas is critical for scaling local observations to regional and global assessments of forest health, carbon stocks, and urban greenery. In an era marked by rapid environmental change, comprehensive mapping efforts are indispensable. Tree crown segmentation is a cornerstone task for these objectives, as the size and structure of a tree crown—shaped by species-specific branching patterns, site conditions, and competition for light—directly influences primary production. A key metric in this context is the crown projection area (CPA), defined as the vertical projection of the crown onto a horizontal plane. Deriving CPA provides essential insights both at the individual tree level—where it provides predictions of diameter, volume, and growth rates—and at the stand level, supporting models of competition and canopy gap dynamics. However, inaccuracies inherent in the segmentation process, particularly when applied at large scales, can lead to biased estimations of these critical tree variables, underscoring the urgent need to address robust validation challenges in large-scale tree crown segmentation. While recent advances in deep learning offer unprecedented capabilities for automated segmentation, their robust validation at scale presents a critical, unaddressed hurdle.

Among the various tree crown variables, crown spread presents unique estimation challenges, both in field measurements and, critically, when derived from remote sensing imagery. Its accurate derivation is not only susceptible to inaccuracies in the initial segmentation process but also relies heavily on the methods employed to extract crown spread from segmented polygons, a process fraught with its own complexities. Selecting an appropriate crown spread calculation method is crucial, especially for irregularly shaped tree crown polygons. Even with optimized calculation methods, direct matches with ground-truth data proved low (e.g., 32% in our experience), although a broader acceptable margin of error (±5 m) captured 89% of cases. These results immediately point to difficulties in establishing precise validation metrics. Furthermore, model performance exhibits size-dependent biases: trees with small crowns (\<6 m) are often overestimated, while large crowns (\>16 m) tend to be underestimated. These discrepancies underscore the inherent limitations of deep learning segmentation models at the extremes of the size spectrum. Critically, the quality of validation data profoundly influences the assessment of crown spread accuracy. For instance, our experience with the local tree inventory ('Baumkataster' Halle/Saale), which provides only broad crown spread ranges (5 m intervals), severely constrains the precision of validation, highlighting the need for more granular and reliable ground-truth datasets.

The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems \[@zheng2024review, @zhao2023systematic\]. Tree crown segmentation, a task central to assessing forest health, carbon stocks, and urban greenery, has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures. Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns, as evidenced by works such as @weinstein2019individual, @khant2025 and @freudenberg2022individual, illustrating the potential for scalable and automated tree mapping. The urgency of climate change and biodiversity loss further amplifies the importance of large-scale tree crown segmentation. Forests serve as critical carbon sinks and biodiversity reservoirs \[@pan2024enduring\], while urban trees provide essential ecosystem services \[@sharma2024urban\]. Large-scale segmentation enables consistent, detailed monitoring of these vital resources, informing sustainable forest management, urban planning, and policy development. The ambition to create national and global inventories, such as those envisioned in @tolan2024very, hinges on the reliability of such foundational segmentation data.

However, while these deep learning models demonstrate impressive performance in controlled environments or with limited-scale datasets, their deployment at expansive spatial extents, such as regional or national scales, exposes a distinct spectrum of challenges that remain largely unaddressed. Foremost among these is validation—a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs. As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, and image quality \[@moussaid2021tree; @cong2022citrus\]. These factors, when coupled with the complexities of real-world landscapes, underscore the need for robust and scalable validation frameworks.

This Perspective builds on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany, addressing federal state-level landscapes. By synthesizing insights from model training, deployment, and validation, we highlight the nuanced challenges associated with assessing segmentation accuracy over expansive and heterogeneous terrains. These challenges are not mere technicalities—they bear significant implications for downstream applications in forestry, conservation, and urban planning. Through this work, we aim to not only highlight these critical validation hurdles but also to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales demanded by global environmental challenges.

## Case Study: Tree Crown Segmentation in Saxony and Saxony-Anhalt, Germany

In this case study, the DeepTrees model \[@khan2025\] was applied in a one-shot prediction approach using pretrained model weights provided by DeepTrees, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). Sachsen, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance.

DOP20 imagery provides detailed spatial information at 20 cm resolution per pixel, capturing visual and near infrared spectral regions essential for accurate tree crown delineation. Using these pretrained weights, the DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in Saxony and 81.4 million in Saxony-Anhalt. The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi), facilitating further research and validation.

This comprehensive segmentation highlights substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients. The case study thus forms a practical foundation for addressing the validation challenges detailed subsequently, emphasizing the necessity for robust validation frameworks to ensure reliable ecological monitoring and informed decision-making based on large-scale automated analyses.

*Figure 1: 6 panels (1 for each state). From left to right \-\> 1\) tiles with date classification 2\) Land use (CORINE) \+ isolines. 3\) the tree segmentations.*

## Challenges

Validating large-scale tree crown segmentation models reveals a web of interrelated challenges that go beyond those encountered in small, controlled datasets. Seasonal phenology is a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes. Models trained on one season often struggle in another, yielding inconsistent segmentation accuracy across the year. For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter. Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed. Incorporating multi-season data during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability \[@moussaid2021tree; @cong2022citrus\].

Spatial heterogeneity of landscapes, including terrain and illumination differences, poses another major hurdle for both segmentation and its validation. An algorithm that segments tree crowns flawlessly in a flat, well-lit park may stumble in a shadow-drenched valley or on a steep mountainside. Variations in ground elevation and slope alter the angle of solar illumination, leading to uneven lighting and shadows that can confuse models. In mountainous or rugged terrains, trees on north-facing slopes might appear darker or partially occluded compared to those on south-facing slopes with direct sun, even if they are the same species and healthy. Such effects result in site-specific performance: models often need fine-tuning or can suffer accuracy drops when moved to new topographies or sensor angles. Weinstein et al. (2020) observed this kind of cross-site performance gap, where a tree detection model trained in one region underperformed when applied to a different region’s imagery without adaptation, underscoring how terrain and context influence outcomes \[@weinstein2020cross\]. For validation, this means that accuracy estimates from one area may not transfer to another – a critical issue when assessments at national or global scales are required.

A further fundamental challenge lies in the scarcity of accurate ground-truth data at scale. Reliable validation hinges on high-quality reference data (the “ground truth”), yet collecting detailed crown delineations over large regions is logistically difficult and expensive. Field surveys can map individual trees with great precision (e.g., measuring trunks and canopy spread on the ground), but doing this over thousands of square kilometers is infeasible. Aerial and satellite imagery provide broader coverage for reference data, but even these often lack the resolution or fidelity to unequivocally label each tree crown for validation purposes. UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth. The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[^\]. This mismatch means that validating a “wall-to-wall” tree map often involves extrapolating from a tiny fraction of ground-referenced trees, introducing uncertainty. Moreover, ground-reference datasets may not capture the full diversity of conditions (species, canopy shapes, management regimes, etc.) present in the larger mapping area, biasing the validation. Expanding ground-truth collection – through automated methods or crowdsourcing – is thus not just a recommendation but a necessity to overcome this validation bottleneck (as we discuss later).

Compounding the issue of limited data is the inconsistency in reference annotations and evaluation metrics. Even when ground-reference data exist, their format can differ – sometimes reference trees are marked by a single GPS point (e.g., trunk location), sometimes by a hand-drawn polygon outlining the crown. This creates a challenge in validation: how do we decide if a predicted crown polygon “matches” a ground-truth point, or how to handle cases where one field-mapped tree corresponds to multiple overlapping crown segments in the image?(Figure ref)  Conversely, field crews might delineate a broad canopy as one crown while an automated model splits it into two segments (or vice versa), especially in dense stands where crowns merge. These ambiguities in one-to-one correspondence make it hard to define what a “correct” segmentation is. Traditional pixel-wise accuracy metrics like Intersection-over-Union (IoU) treat segmentation purely as an image overlap problem, which may not reflect the ecological reality of counting individual trees. IoU penalizes differences in shape or area but doesn’t account for whether the count of tree objects is correct. In an extreme case, a model could slightly over-segment every tree (splitting each true crown into two smaller polygons) and still achieve a reasonable IoU, despite doubling the perceived tree count – a significant error for applications. On the other hand, object-centric metrics such as panoptic segmentation quality attempt to consider both detection and delineation of objects \[@kirillov2019panopticsegmentation\]. Panoptic metrics combine aspects of object detection (was each tree detected?) with segmentation quality (was each crown correctly outlined?), which can be more appropriate for tree mapping. However, even these require well-defined ground-truth objects to compare against. When the ground truth itself is inconsistent (e.g., how to count a clumped cluster of stems with overlapping crowns), validation metrics struggle to fully capture model performance. The choice of evaluation metric thus becomes non-trivial: depending on whether one prioritizes exact crown shape, tree count, or canopy cover, the “best” metric may differ. Establishing consensus on evaluation protocols is part of the challenge – without it, different studies may report accuracy in incompatible ways.

There is also the issue of scale and resolution in validation reporting. A model’s accuracy can appear to vary depending on the spatial scale at which it’s evaluated. For instance, a segmentation model might achieve high overall accuracy when averaged over an entire region, yet if one zooms into a small test area (say a single city park or forest stand), the error rate might be much higher or lower. This can happen if errors are not evenly distributed: the model could perform very well in one type of landscape (e.g., neat urban street trees) and poorly in another (dense natural forest), and a coarse regional average could mask these extremes. Consequently, a user working on a local conservation project might experience worse performance than the “headline” accuracy suggests, because that headline number was diluted by many easier cases elsewhere. Ensuring that validation is robust across scales is tricky – one must balance broad coverage with local detail. It calls for multi-scale validation approaches, where accuracy is reported at multiple grain sizes or stratified by landscape type. Highlighting this, one could imagine a figure plotting model accuracy as a function of spatial extent or across different habitat types, illustrating how performance can drop off in specific challenging subsets despite looking good overall. This emphasizes that heterogeneous performance is itself a challenge to acknowledge in validation: we need methods to detect where and why a model fails, not just an aggregate score.

To summarize, validating large-scale tree segmentation models is far from a straightforward task. Seasonal changes, diverse terrain and illumination, limited and inconsistent ground truth, ambiguous evaluation criteria, and scale-dependent performance all intertwine to create a demanding setting. These challenges are not just academic – they directly impact how much trust we can place in AI-generated tree maps for real-world decisions in forestry, ecology, and urban planning. Recognizing these pain points is the first step; the next is devising strategies to overcome them. We now turn to several recommendations aimed at improving both models and validation practices, with an eye toward bridging the gap between controlled experiments and the complexity of continental-scale deployments.

*Table 1: Potential validation challenges and corresponding mitigation strategies. This table could list key challenges in validating large-scale tree segmentations alongside proposed approaches to address them. For example: (Challenge) Phenological variation between seasons – (Mitigation) use multi-temporal training data and season-specific validation sets; (Challenge) Heterogeneous terrain and shadowing – (Mitigation) integrate digital elevation models and terrain-aware modeling; (Challenge) Scarcity of ground-truth annotations – (Mitigation) leverage semi-automated labeling, crowdsourcing, and active learning to expand validation data; (Challenge) Ambiguity in crown delineation – (Mitigation) adopt object-based accuracy metrics and consensus protocols for what counts as a single tree; (Challenge) Scale-dependent performance – (Mitigation) evaluate models at multiple spatial scales and stratify results by landscape type. This overview would visually reinforce the narrative that each challenge has an identifiable path forward, linking the “Challenges” and “Recommendations” sections.*

*![][image1]*  
*Figure 2: Incomplete*

## Recommendations

Addressing the above challenges requires a multi-pronged approach, combining improvements in modeling techniques with innovations in validation methodology and data collection. We outline here several complementary strategies that, together, can bolster the reliability of large-scale tree segmentation efforts. These recommendations emphasize building more robust models through advanced training paradigms, as well as creating better frameworks to evaluate and support those models in real-world conditions. The overarching goal is to ensure that deep learning advancements in tree mapping translate into trustworthy, actionable insights at regional to global scales.

**Leverage self-supervised learning for robust representations.** A key step towards more generalizable segmentation models is tapping into unannotated data via self-supervised learning (SSL). Unlike traditional supervised training which is bottlenecked by limited labeled examples, SSL allows models to learn from the abundant pool of unlabeled remote sensing images – for example, by predicting missing pieces of an image or distinguishing augmented views of the same scene. By pre-training a model on large geospatial datasets without any human-provided labels, the model can absorb intrinsic patterns of the landscape: textures, shapes, seasonal changes, and other context that are common across images. These rich foundational representations can then be fine-tuned for tree crown segmentation with far fewer annotated samples than would otherwise be needed. Recent efforts like PhilEO Bench demonstrate the promise of this approach, evaluating geospatial foundation models that were pre-trained using SSL techniques \[@fibaek2024phileo\]. The results show improved performance on a range of tasks (e.g. building footprint extraction, road mapping) after such pre-training, compared to models trained from scratch. In the context of trees, a model with an SSL-pretrained backbone may already “know” about basic vegetation structures, shadows, and seasonal appearances, making it more adept at delineating crowns under varied conditions. For instance, a foundation model trained on year-round satellite imagery might implicitly understand the difference between a leafless oak in winter and the same oak in summer, and thus require only a light fine-tuning to accurately segment each. Mendieta et al. (2023) take this further by continually pre-training on new data distributions (a form of continual learning), which helped build a Geospatial Foundation Model (GFM) that excelled across multiple remote sensing tasks \[@mendieta2023gfm\]. Such continual SSL training could allow segmentation models to keep improving as more unlabeled data (e.g., new satellite images over time or from new regions) become available, staying up-to-date with changes in landscapes and sensor characteristics. Exploiting SSL is a powerful way to tackle the twin issues of limited labels and dataset bias, yielding a model that is better equipped for the diverse scenarios encountered in large-scale mapping.

>>>>

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author:
  - name: Taimur Khan
    corresponding: true 
    affiliations:
      - name: Helmholtz Centre for Environmental Research (UFZ)
        address: Theodor-Lieser-Str. 4, 06120 Halle (Saale)
        country: Germany
        url: \{taimur.khan\}\@ufz.de
        id: south
    url: https://www.ufz.de/index.php?en=49404
  - name: Calvin Cordozar Broadus Jr.
    affiliations:
      - name: Fictional West Coast University
        address: Long Beach CA 90840
        country: USA
        url: ccb@fwcu.fake
      - name: Secondary European Affiliation
        address: Tiergartenstr. 17
        city: Heidelberg
        country: Germany
        postal-code: 69121 
        url: lncs@springer.com
  - name: Antwan André Patton
    orcid: '0000-1111-2222-3333'
    affiliations:
      - ref: south 
author-running: T. Khan et al.
abstract: |
 Recent progress in deep learning has greatly improved the ability to extract detailed information from remote sensing images, such as identifying and outlining individual tree crowns. However, as these models are scaled up for use across large geographic areas—beyond individual forests or research plots—new challenges begin to emerge. One of the main hurdles is validating how accurately these models detect and segment tree crowns. Even small errors in these predictions can affect decisions in forestry management and urban planning, limiting the real-world usefulness of such models. Our experience working with large-scale applications across regions in Germany highlights how factors like tree size, seasonal variations, shadow effects, and the shape of training data annotations influence model performance. This underlines the importance of thoughtful selection and design of training datasets. We discuss these challenges in validation and emphasize the need for careful model evaluation when working at broader spatial scales.
keywords: [First Keyword, Second Keyword, Another Keyword.]
---

# Introduction

The urgency of climate change and biodiversity loss significantly heightens the importance of large-scale tree crown segmentation. Forests act as crucial carbon sinks and biodiversity reservoirs [@pan2024enduring], while urban trees offer vital ecosystem services[@sharma2024urban]. To generate comprehensive assessments of forest health, carbon stocks, and urban greenery, we need to accurately map tree crowns over vast areas, including trees both within forests and outside of them. These trees outside forests, often found in agricultural or urban settings, must be part of our inventories, making their detection and mapping a key priority for scaling local observations to a global context. Large-scale segmentation enables consistent, detailed monitoring of these vital resources, informing sustainable forest management, urban planning, and policy development. The ambition to create national and global inventories, such as those envisioned in @tolan2024very, hinges on the reliability of such foundational segmentation data. In an era marked by rapid environmental change, comprehensive mapping efforts are indispensable. Tree crown segmentation is a cornerstone task for these objectives, as the size and structure of a tree crown—shaped by species-specific branching patterns, site conditions, and competition for light—directly influences primary production.

While recent advances in deep learning offer unprecedented capabilities for automated segmentation, their robust validation at scale presents a critical, unaddressed hurdle.

The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems [@zheng2024review; @zhao2023systematic]. Tree crown segmentation has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures. Latter was used by [@JIANG2025103114], for single tree detection. Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns [@weinstein2019individual; @khant2025deeptrees; @freudenberg2022individual], illustrating the potential for scalable and automated tree mapping.

However, while these deep learning models demonstrate impressive performance in controlled environments or with limited-scale datasets, their deployment at expansive spatial extents, such as regional or national scales, exposes a distinct spectrum of challenges that remain largely unaddressed. Foremost among these is validation—a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs. As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, vitality conditions, and image quality [@moussaid2021tree; @cong2022citrus], a fact that directly reflects the profound complexity of how tree crowns appear in real-world conditions. To enable scalable validation frameworks, it is therefore necessary to accurately characterize and account for this complexity.

A key metric in this context is the crown projection area (CPA), defined as the vertical projection of the crown onto a horizontal plane. Deriving CPA provides essential insights both at the individual tree level—where it provides predictions of diameter, volume, and growth rates—and at the stand level, supporting models of competition and canopy gap dynamics [@shimano1997analysis; @krajicek1961crown]. However, inaccuracies inherent in the segmentation process, particularly when applied at large scales, can lead to biased estimations of these critical tree variables, underscoring the urgent need to address robust validation challenges in large-scale tree crown segmentation.

Among the available crown metrics, crown spread is particularly valuable as an independent variable for validating segmentation results, as it captures essential information about tree size and structure. However, accurate estimation of crown spread is complex in both field and remote sensing contexts. Errors can arise not only during the segmentation process but also from the methods used to derive crown spread from segmented polygons, especially for irregularly shaped crowns. The choice of calculation method strongly influences the quality of crown spread estimates and, consequently, the reliability of validation metrics. Furthermore, the accuracy of validation is fundamentally dependent on the quality of reference data; coarse or inconsistent ground-truth datasets can substantially limit the precision of assessment.

Building on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany at the federal‐state level, this paper (1) identifies and characterizes the key challenges involved in assessing segmentation accuracy over expansive, heterogeneous terrains––including variability in canopy structure, imaging conditions, and regional ecological gradients––and (2) provides concrete methodological recommendations to address these challenges. By coupling a rigorous evaluation of terrain‐specific segmentation limitations with these targeted recommendations, we aim to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales required by global environmental challenges.

# Case Study: Tree Crown Segmentation in Saxony and Saxony-Anhalt, Germany

To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3). To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3).

![]("./figures/figure1.png")

# Challenges

Validating large-scale tree crown segmentation models reveals a web of interrelated challenges. Among these challenges is phenology, a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes. Models trained on a single phenological period often struggle in another, yielding inconsistent segmentation accuracy across the year [@sapkota2024; @liu2024]. For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter. Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed. Incorporating multi-season data including the key phenological stages: leaf emergence, flowering, fruiting, autumn coloring and leaf fall, during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability [@moussaid2021tree; @cong2022citrus].

Spatial and illumination heterogeneity of landscapes, driven by terrain and illumination differences, poses a major hurdle for both segmentation and its validation. An algorithm that segments tree crowns flawlessly in a flat, well-lit park may strugglein a shadow-drenched valley or on a steep mountainside. Variations in ground elevation and slope alter the angle of solar illumination, leading to uneven lighting and shadows that can confuse models. In mountainous or rugged terrains, trees on north-facing slopes might appear darker or partially occluded compared to those on south-facing slopes with direct sun, even if they are the same species and healthy. Such effects result in site-specific performance: accuracy drops when moving to new topographies or sensor angles. Weinstein et al. (2020) observed this kind of cross-site performance gap, where a tree detection model trained in one region underperformed when applied to a different region’s imagery without adaptation, underscoring how terrain and context influence outcomes [@weinstein2020cross]. Fine-tuning can increase accuracy, but requires additional data and computational resources. For validation, this means that accuracy estimates from one area may not transfer to another – a critical issue when assessments at national or global scales are required.

A further fundamental challenge lies in the scarcity of accurate ground-truth data at scale. Reliable validation hinges on high-quality reference data (the "ground truth"), yet collecting detailed crown delineations over large regions is logistically difficult and expensive. For instance, field surveys can map individual trees with great precision, providing highly accurate tree position information (e.g., through GPS measurements of trunk location or detailed canopy spread on the ground) [@steier2024]. However, doing this over thousands of square kilometers is infeasible. Conversely, while aerial and satellite imagery offer broad coverage, even at high resolutions, they present significant challenges for unequivocally labeling each tree crown for validation purposes. This is due to factors like canopy overlap, varied lighting conditions, complex tree morphologies, and the sheer scale of the areas to be annotated, which introduce ambiguities and make consistent manual delineation impractical or prohibitively expensive across large regions. UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth. The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[^\]. This mismatch means that validating a “wall-to-wall” tree map often involves extrapolating from a tiny fraction of ground-referenced trees, introducing uncertainty. Moreover, ground-reference datasets may not capture the full diversity of conditions (species, canopy shapes, management regimes, etc.) present in the larger mapping area, biasing the validation. Expanding ground-truth collection – through automated methods or crowdsourcing – is thus not just a recommendation but a necessity to overcome this validation bottleneck (as we discuss later).

Compounding the issue of limited data is the inconsistency in reference annotations and evaluation metrics. Even when ground-reference data exist, their format can differ – sometimes reference trees are marked by a single GPS point (e.g., trunk location), sometimes by a hand-drawn polygon outlining the single crown. This creates a challenge in validation: how do we decide if a predicted crown polygon “matches” a ground-truth point, or how to handle cases where one field-mapped tree corresponds to multiple overlapping crown segments in the image? (Figure 2c) Conversely, field crews might delineate a broad canopy as one crown while an automated model splits it into two segments (or vice versa), especially in dense stands where crowns merge. These ambiguities in one-to-one correspondence make it hard to define what a “correct” segmentation is. Traditional pixel-wise accuracy metrics like Intersection-over-Union (IoU) treat segmentation purely as an image overlap problem, which may not reflect the ecological reality of counting individual trees. IoU penalizes differences in shape or area but doesn’t account for whether the count of tree objects is correct. In an extreme case, a model could slightly over-segment every tree (splitting each true crown into two smaller polygons) and still achieve a reasonable IoU, despite doubling the perceived tree count – a significant error for applications. On the other hand, object-centric metrics such as panoptic segmentation quality attempt to consider both detection and delineation of objects [@kirillov2019]. Panoptic metrics combine aspects of object detection (was each tree detected?) with segmentation quality (was each crown correctly outlined?), which can be more appropriate for tree mapping. However, even these require well-defined ground-truth objects to compare against. When the ground truth itself is inconsistent (e.g., how to count a clumped cluster of stems with overlapping crowns), validation metrics struggle to fully capture model performance. The choice of evaluation metric thus becomes non-trivial: depending on whether one prioritizes exact crown shape, tree count, or canopy cover, the “best” metric may differ. Establishing consensus on evaluation protocols is part of the challenge – without it, different studies may report accuracy in incompatible ways.

There is also the issue of scale and resolution in validation reporting. A model’s accuracy can appear to vary depending on the spatial scale at which it is evaluated. For instance, a segmentation model might achieve high overall accuracy when averaged over an entire large region, yet if one zooms into a small test area (say a single city park or forest stand), the error rate might be much higher or lower. This can happen if errors are not evenly distributed: the model could perform very well in one type of landscape (e.g., neat urban street trees) and poorly in another (dense natural forest), and a coarse regional average could mask these extremes. Consequently, a user working on a local conservation project might experience worse performance than the “headline” accuracy suggests, because that headline number was diluted by many easier cases elsewhere. Ensuring that validation is robust across scales is tricky – one must balance broad coverage with local detail. It calls for multi-scale validation approaches, where accuracy is reported at multiple grain sizes or stratified by landscape or habitat  type.

The above challenges reflect the multifaceted difficulties of validating tree crown segmentation at scale. For an at-a-glance overview of these issues and recommendations, see Table 1 in the Appendix.

![]("./figures/figure2.png")

*Figure 2:* Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.

# Recommendations

Validating large-scale tree segmentation models is challenging due to seasonal variability, diverse terrain and illumination conditions, inconsistent ground truth data, ambiguous evaluation criteria, and scale-dependent performance. These factors significantly affect the reliability and practical value of AI-generated tree maps for forestry, ecology, and urban planning.

Addressing these challenges necessitates a multi-faceted approach that integrates advanced modeling techniques, improved validation methods, and robust data collection. Recommended strategies include enhancing model robustness through sophisticated training approaches and developing standardized evaluation frameworks to assess real-world performance accurately. These efforts aim to ensure that advances in deep learning yield reliable, actionable insights for large-scale tree mapping.

**Leveraging self-supervised learning (SSL)** enhances the generalizability of segmentation models by utilizing abundant unlabeled remote sensing data. Unlike traditional supervised methods constrained by limited annotations, SSL enables models to derive meaningful representations through tasks such as predicting missing image parts or distinguishing augmented views of scenes. Pre-training models with extensive unlabeled geospatial imagery allows them to internalize intrinsic landscape patterns—textures, shapes, and seasonal variations—which can be efficiently fine-tuned with fewer labeled samples for specific tasks like tree crown segmentation. Recent research underscores the effectiveness of SSL-based geospatial foundation models. PhilEO Bench demonstrated improved performance across multiple remote sensing applications, such as building footprint extraction and road mapping, compared to training models from scratch [@fibaek2024phileo]. SSL-trained models inherently recognize basic vegetation structures, shadows, and seasonal dynamics, facilitating accurate segmentation under varying conditions. For example, a model trained on seasonal satellite imagery implicitly distinguishes between a tree’s winter and summer appearances, significantly reducing the required fine-tuning. Expanding this approach, Mendieta et al. (2023) employed continual SSL training, integrating new data distributions to develop a robust Geospatial Foundation Model (GFM) proficient across diverse remote sensing tasks [@mendieta2023gfm]. This continual learning paradigm ensures segmentation models remain current with evolving landscapes and sensor technologies. Thus, SSL effectively addresses challenges associated with limited labeled data and dataset biases, improving adaptability and performance of large-scale segmentation models.

**Integrating multi-view and multi-temporal data** enhances model consistency by training with diverse imagery of the same trees captured from varying angles, sensors, or times. Multi-view data encompass multi-angle (nadir and oblique aerial images), multi-platform (satellite and drone imagery), and multi-temporal (images from different seasons or years) perspectives, enabling models to learn robust invariances for reliable segmentation under varying conditions. Combining leaf-on and leaf-off images, compels the model to recognize structural features beyond mere greenness, thus improving seasonal generalization. Self-supervised learning (SSL) techniques, such as masked autoencoders and contrastive learning, effectively utilize multi-view datasets by encouraging models to generate consistent representations across different views without manual labels. Studies utilizing masked image modeling and contrastive SSL on multi-view satellite imagery have demonstrated significant performance improvements in segmentation and detection tasks [@mukkavilli2023prithvi; @fibaek2024phileo]. Employing multi-view consistency training directly mitigates challenges associated with phenological and illumination variations, thereby enhancing model robustness and facilitating validation. For example, discrepancies in model predictions between leaf-on and leaf-off imagery could flag areas requiring further examination. Thus, multi-view and multi-temporal training provide models with contextual understanding of the dynamic and three-dimensional nature of trees, stabilizing segmentation results across real-world variability.
